---
title: "assigment2.0"
output: html_document
date: "2023-11-02"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(dplyr)
library(stringr)
library(ggplot2)
library(gridExtra)
library(rlang)
library(tidyr)
```

# Import files

```{r}
covariates <- read_csv('covariates.csv')
covariates <- covariates[1:6] %>% filter(!is.na(PatientID))
biomarkers <- read_csv('biomarkers.csv') 
biomarkers <- biomarkers %>% filter(!is.na(Biomarker))
```

# Clean data

## Split Biomarker into PatientID and Timepoint
```{r}
biomarkers <- biomarkers %>% 
  separate(Biomarker, into = c("PatientID", "Timepoint"), sep = '-')
```

## Dummy code sex and smoking status 
Non-smokers = 0, Non smokers = 1
Males = 0, Females = 1
```{r}
covariates <- covariates %>%
  mutate(Smoker = recode(`Smoker (1=yes, 2=no)`, `2` = 0), 
         Sex = recode(`Sex (1=male, 2=female)`, `1` = '0', `2`= '1')) 

#Remove original Sex and Smoker variables and turn PatientID into character
covariates <- covariates[-c(3,4)] %>% 
  mutate(PatientID = as.character(PatientID))
```

## Join datasets
Only want observations that have data for both biomarker levels and covariates
```{r}
meta <- biomarkers %>% 
  inner_join(covariates, join_by(PatientID))
```

## Get rid of '-' in column names 
```{r}
colnames(meta) <- gsub("-", "", colnames(meta))
```


# PART 1: HYPOTHESIS TESTING ##########################################

Question of choice: Do the biomarker levels (at inclusion) for patients with high VAS scores differ from those with low VAS scores at inclusion?

## Filter observations to only the measurements at inclusion

```{r}
meta.inclusion <- meta %>% 
  filter(Timepoint == "0weeks") %>% 
  mutate(VAS.category = if_else(`VASatinclusion` > 5, "High", "Low"))
```

## Visually inspect distributions of biomarkers 

```{r}
# Write function to make life easier
plot_biomarker_histogram <- function(df, biomarker) {
  n_high <- nrow(df %>% filter(VAS.category == 'High'))
  n_low <- nrow(df %>% filter(VAS.category == 'Low'))
  
  p <- ggplot(df, aes(.data[[biomarker]])) +
    geom_histogram(aes(fill = VAS.category), alpha = 0.5, bins = 30) +
    facet_grid(. ~ VAS.category) +
    ggtitle(paste(biomarker, "\nHigh VAS score: ", n_high, " observations\nLow VAS score: ", n_low, " observations")) +
    xlab(paste(biomarker, "Levels")) +
    ylab("Frequency") +
    theme_minimal()
  
  print(p)
}

biomarker_list <- c("IL8", "VEGFA", "OPG", "TGFbeta1", "IL6", "CXCL9", "CXCL1", "IL18", "CSF1")

for(biomarker in biomarker_list) {
  plot_biomarker_histogram(meta.inclusion, biomarker)
}

```

## Check whether data follows a normal distribution

### Normality (Q-Q) plots

```{r}
# Function to create normality plot
plot_normality_plot <- function(df, biomarker) {
  np <- ggplot(df, aes(sample = .data[[biomarker]])) +
    geom_qq() +
    geom_qq_line() +
    facet_grid(. ~ VAS.category) +
    ggtitle(paste("Q-Q Plot for", biomarker)) +
    xlab(paste(biomarker, "Theoretical Quantiles")) +
    ylab(paste(biomarker, "Sample Quantiles"))
  
  print(np)
  
}

# Loop through each biomarker
for(biomarker in biomarker_list) {
  plot_normality_plot(meta.inclusion, biomarker)
}

```

### Shapiro-Wilk test

```{r}
# Create data frame to store results
test_results_df <- data.frame(
  Biomarker = character(),
  Category = character(),
  W = numeric(),
  p_value = numeric(),
  Deviation_from_normality = character(),
  stringsAsFactors = FALSE
)

# Loop through Shapiro-Wilk test for each biomarker and each category
for (biomarker in biomarker_list) {
  for (category in c("High", "Low")) {
    data_subset <- meta.inclusion %>% filter(VAS.category == category) %>% pull(!!sym(biomarker))
    test <- shapiro.test(data_subset)
    
    deviation <- ifelse(test$p.value < 0.05, "Yes", "No")
    
    result <- data.frame(
      Biomarker = biomarker,
      Category = paste0("VAS", tolower(category)),
      W = test$statistic,
      p_value = test$p.value,
      Deviation_from_normality = deviation,
      stringsAsFactors = FALSE
    )
    
    test_results_df <- rbind(test_results_df, result)
  }
}

# Print results
print(test_results_df)

# Plot graph of p-values from Shapiro-Wilk test for each biomarker to quickly visualise results of shapiro-wilk test
sw.plot <- ggplot(test_results_df, aes(x = Biomarker, y = p_value, color = Category)) +
  geom_point(size = 3) + 
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "red") +
  theme_minimal() +
  labs(title = "Plot of the p-values from Shapiro-Wilk Test for each biomarker",
       y = "p-value",
       x = "Biomarker") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.title = element_blank()) +
  scale_y_continuous(limits = c(0, 1))

# Display the plot
print(sw.plot)
```

## Perform Student's T-test
Parametric test. Do not assume equal variances between high and low VAS patient groups, and so use Welch's t-test

```{r}
# Create df
results_df <- data.frame(
  Biomarker = character(),
  t_value = numeric(),
  df = numeric(),
  p_value = numeric(),
  Significance = character(),
  CI_lower = numeric(),
  CI_upper = numeric()
)

# Loop through each biomarker to perform the Welch's t-test
for (biomarker in biomarker_list) {
  
  # Data for high VAS
  data_VAShigh <- meta.inclusion %>% filter(VAS.category == 'High') %>% pull(biomarker)
  
  # Data for low VAS
  data_VASlow <- meta.inclusion %>% filter(VAS.category == 'Low') %>% pull(biomarker)
  
  # Perform Welch's t-test
  test_result <- t.test(data_VAShigh, data_VASlow, var.equal = FALSE)
  
  # Check for significance at 95% confidence level
  significance <- ifelse(test_result$p.value < 0.05, "Significant", "Not Significant")
  
  # Store the results in the data frame
  results_df <- rbind(results_df, data.frame(
    Biomarker = biomarker,
    t_value = test_result$statistic,
    df = test_result$parameter,
    p_value = test_result$p.value,
    Significance = significance,
    CI_lower = test_result$conf.int[1],
    CI_upper = test_result$conf.int[2]
  ))
}

# Print the results table
print(results_df)

```

## Perform sensitivity analysis with Mann-Whitney (Wilcoxin rank-sum) non-parametric test

```{r}
# Create an empty data frame to store the results
results_df.mw <- data.frame(
  Biomarker = character(),
  U_value = numeric(),
  p_value = numeric(),
  CI_lower = numeric(),
  CI_upper = numeric(),
  Significance = character()
)

# Loop through each biomarker
for (biomarker in biomarker_list) {
  
  # Data for high VAS
  data_VAShigh <- meta.inclusion %>% filter(VAS.category == 'High') %>% pull(biomarker)
  
  # Data for low VAS
  data_VASlow <- meta.inclusion %>% filter(VAS.category == 'Low') %>% pull(biomarker)
  
  # Perform Mann-Whitney U Test
  test_result.mw <- wilcox.test(data_VAShigh, data_VASlow, alternative = "two.sided", conf.int = TRUE) # Enable confidence interval
  
  # Check for significance at 95% confidence level
  significance <- ifelse(test_result.mw$p.value < 0.05, "Significant", "Not Significant")
  
  # Store the results in the data frame
  results_df.mw <- rbind(results_df.mw, data.frame(
    Biomarker = biomarker,
    U_value = test_result.mw$statistic,
    p_value = test_result.mw$p.value,
    CI_lower = test_result.mw$conf.int[1],
    CI_upper = test_result.mw$conf.int[2],
    Significance = significance
  ))
}

# Print the results table
print(results_df.mw)
```

## Welch's two-tailed t-test with Bonferroni Correction

```{r}
# Create an empty data frame to store the results
results_df.bc <- data.frame(
  Biomarker = character(),
  t_value = numeric(),
  df = numeric(),
  p_value = numeric(),
  Significance = character(),
  CI_lower = numeric(),
  CI_upper = numeric()
)

# Number of tests to be performed
num_tests <- length(biomarker_list)

# Adjusted significance level (alpha) based on Bonferroni correction
adjusted_alpha <- 0.05 / num_tests

# Loop through each biomarker to perform the Welch's t-test
for (biomarker in biomarker_list) {
  
  # Data for high VAS
  data_VAShigh <- meta.inclusion %>% filter(VAS.category == 'High') %>% pull(biomarker)
  
  # Data for low VAS
  data_VASlow <- meta.inclusion %>% filter(VAS.category == 'Low') %>% pull(biomarker)
  
  # Perform Welch's t-test
  test_result.bc <- t.test(data_VAShigh, data_VASlow, var.equal = FALSE, conf.level = (1-adjusted_alpha))
  
  # Check for significance using the adjusted alpha level
  significance <- ifelse(test_result.bc$p.value < adjusted_alpha, "Significant", "Not Significant")
  
  # Store the results in the data frame
  results_df.bc <- rbind(results_df.bc, data.frame(
    Biomarker = biomarker,
    t_value = test_result.bc$statistic,
    df = test_result.bc$parameter,
    p_value = test_result.bc$p.value,
    Significance = significance,
    CI_lower = test_result.bc$conf.int[1],
    CI_upper = test_result.bc$conf.int[2]
  ))
}

# Print the results table
print(results_df.bc)
```

# Regression Modelling #############

## Split data into train and test observations/set
Already filtered the data-frame to only have observations for biomarker levels at inclusion in previous section. Please note that from here on, the code used will produce the same values given in my report due to the use of the sample_frac() function, which generates a random sample (with a random seed) each time - so the data will be split differently if you try run it to check my predictions. I have confirmed that this is okay with Course Organiser. 

```{r}
# Remove any observations with NA values
meta.inclusion.noNA <- meta.inclusion[rowSums(is.na(meta.inclusion)) == 0, ]

# Randomly sample 80% of the data for the training sets
train_data <- meta.inclusion.noNA %>% sample_frac(.8)

# Use the remaining 20% for the test set
test_data <- anti_join(meta.inclusion.noNA, train_data)
```

## Fit model

```{r}
m1 <- lm(Vas12months ~ IL8 + VEGFA + OPG + TGFbeta1 + IL6 + CXCL9 + CXCL1 + IL18 + CSF1 + Age + Smoker + Sex + VASatinclusion, train_data)
summary(m1)
confint(m1, level = 0.95)
```

### Residual plots

```{r}
# Histogram
hist(residuals(m1), 
     main = "Histogram of Residuals for training dataset", 
     xlab = "Residual Value", 
     ylab = "Frequency", 
     xlim=c(-6, 6), 
     ylim=c(0,10),
     breaks=30) 


```
Displays a slight right skew


```{r}
# Residuals against predicted values
plot(predict(m1), residuals(m1), 
     xlab = "Predicted 12-month VAS Scores", 
     ylab = "Residuals", 
     main = "Residuals vs. Predicted 12-month VAS Scores", 
     xlim = c(0, 10), 
     ylim = c(-7,7))
```
Residuals look almost completely random here

## Make predictions of 12-month VAS scores for the remaining 20% of patients 

```{r}
#Make predictions
predicted_values.test <- predict(m1, test_data, interval = 'prediction', level = 0.95)
```

### Compare to observed values
```{r}
# Bind the Vas12months column from test_data with the predicted_values.test data-frame
combined_prediction_data <- cbind(test_data$Vas12months, predicted_values.test)

# Rename the first column 
colnames(combined_prediction_data)[1] <- "observed"

# Directly checking if 'Vas12months' is outside of the predicted interval
combined_prediction_data <- as.data.frame(combined_prediction_data)

combined_prediction_data <- combined_prediction_data %>% 
  mutate(outside_PI = ifelse(observed < lwr | observed > upr, "Yes", "No"))

# Print the combined data
print(combined_prediction_data)
```

### Plot predictions against actual values for the test set

```{r}
# Scatter plot of observed vs. predicted values
plot(test_data$Vas12months, predicted_values.test[,1], main="Observed vs. Predicted 12-month VAS scores for test data",
     xlab="Observed 12-month VAS score", ylab="Predicted 12-month VAS score", pch=19, col="green",
     xlim=c(0, 10), ylim=c(0, 10))

# Add 45-degree line
abline(a=0, b=1, col="pink", lwd=2)

# Add a legend
legend("topleft", legend=c("Observed vs. Predicted", "Perfect Prediction"),
       col=c("green", "pink"), pch=c(19, NA), lty=c(NA, 1), lwd=c(NA, 2))

```

### Compare to predictions made for the train set

```{r}
# Get predicted values from the model for the training data
predicted_values <- predict(m1, train_data)

# Scatter plot of observed vs. predicted values
plot(train_data$Vas12months, predicted_values, main="Observed vs. Predicted 12-month VAS scores for training data",
     xlab="Observed 12-month VAS score", ylab="Predicted 12-month VAS score", pch=19, col="blue", 
     xlim=c(0, 10), ylim=c(0, 10))

# Add 45-degree line
abline(a=0, b=1, col="red", lwd=2)

# Add a legend
legend("topleft", legend=c("Observed vs. Predicted", "Perfect Prediction"),
       col=c("blue", "red"), pch=c(19, NA), lty=c(NA, 1), lwd=c(NA, 2))

```
